What We Eval in the Shadows

### 2021-08-10 paper/main.tex
Add formula resolution

\mypara{Code Obfuscation} Some packages use \eval to bypass the restrictions
imposed by the \code{R CMD CHECK} tool. This tool enforces some well-formedness
rules on a submitted package before accepting it for inclusion in CRAN. Some
static checks ensure that the package code does not use certain restricted
functions. If package authors cannot avoid using these functions, they use \eval
to obfuscate the code, bypassing the checks performed by the tool. An example of
this use is the \code{scalaEnsure} function from the \code{aibd} package. It
mutates the variable \code{s} in the package environment. However, the package
environment in R is locked by default, which prevents bindings from being
mutated. Environment can be unlocked using \code{unlockBinding}, but its use is
flagged by the well-formedness checks. To work around this limitation, the
author passes the expression to unlock the package environment as a string to
\code{parse} which is then evaluated by \code{eval}.

\begin{figure}[h]
\begin{lstlisting}
scalaEnsure <- function() {
    ...
    env <- parent.env(environment())
    eval(parse(text=paste0('unlockBinding("s", env)')))
    assign("s", s, envir=env)
    lockBinding("s", env)
}
\end{lstlisting}
\caption{aibd::scalaEnsure}
\end{figure}

@@

\mypara{Attributes}
Package \code{nlme} uses \eval to evaluate expressions attached as attributes to
its objects. Consider the definition of the \code{formula.pdMat} function below.
It extracts the \code{"formula"} attribute attached to the \code{pdMat} object
and evaluates it.

>>>>

\mypara{Formula Resolution} Many packages use \code{eval} to evaluate formula
expressions. Consider the definition of the \code{formula.pdMat} function from
\code{nlme} package.

It extracts the \code{"formula"} attribute attached to the \code{pdMat} object
and evaluates it.

\mypara{Code Obfuscation} Some packages use \eval to bypass the restrictions
imposed by the \code{R CMD CHECK} tool. This tool enforces some well-formedness
rules on a submitted package before accepting it for inclusion in CRAN. Some
static checks ensure that the package code does not use certain restricted
functions. If package authors cannot avoid using these functions, they use \eval
to obfuscate the code, bypassing the checks performed by the tool. An example of
this use is the \code{scalaEnsure} function from the \code{aibd} package.
\begin{figure}[h]
\begin{lstlisting}
scalaEnsure <- function() {
    ...
    env <- parent.env(environment())
    eval(parse(text=paste0('unlockBinding("s", env)')))
    assign("s", s, envir=env)
    lockBinding("s", env)
}
\end{lstlisting}
  \caption{aibd::scalaEnsure}
\end{figure}

This function mutates the variable \code{s} in the package environment. However,
the package environment in R is locked by default, which prevents bindings from
being mutated. Environment can be unlocked using \code{unlockBinding}, but its
use is flagged by the well-formedness checks. To work around this limitation,
the author passes the expression to unlock the package environment as a string
to \code{parse} which is then evaluated by \code{eval}.

@@
